---
title: "Logistic Regression"
output: html_document
---

# Reading the data
This week we will learn about logistic regression. Before analysing the data I have been done some wrangeling. It wasn't as diffucult as week before. Now the data is ready to be analyzed. The following changes have been made: the variables not used for joining the two data have been combined by averaging. The variable 'alc_use' is the average of 'Dalc' and 'Walc'and 'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise.
```{r}
alc <- read.csv("alc2014.csv", header = TRUE, sep = ",")

#Looking at the structure

dim(alc)
colnames(alc)
summary(alc)
```
The modified data consists of 382 observations of 35 variables as seen above. The data tells students' achievements in secondary education of two Portuguese schools. This data was collected by questionaires and school reports. Originally we had two datasets regarding the performance in two distinct subjects: Mathematics (called mat) and Portuguese language (called por). If you want, you can find more information about the original dataset here: https://archive.ics.uci.edu/ml/datasets/Student+Performance.

# The purpose of my analysis is to study the relationships between high/low alcohol consumption and some of the other variables in the data. 
To do this I choosed 4 variables in the data. And for each of them I had present my own personal hypothesis of their relationship with alcohol consumption. I chose the variables failure, absences, famrel and higher.
1. Variable "failure" tells about the number of past class failures which might be linked to higher alcohol consumption. When you think with common sense these two variables could have correlation, which is why I chose them.
2. The second chosen variable is "absences". The variables measures school absences. Again when you think with common sense it could be explainde that those who drink more alchol have more school absences. Thus is why I want to see is there correlation or not.
3. The third chosen variable is "famrel". It means the quality of family relationships. It is commonly known that lower social status is connected to more alchol consumption. Now we will find out is there correlation between these two variables in this data.
4. The fourth chosen variable is "higher". This variable tells if the student wants to achieve higher education. It can be argued that those who study more and harder, have lesser time for partying and for instance alchol use. So we will find out it those who want a higher education consume alcohol less.

# Next wer are going to numerically and graphically explore the distributions of my chosen variables and their relationships with alcohol consumption. We are going to use for example cross-tabulations, bar plots and box plots. 

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

```
# First we explore the relationship between alcohol consumption and number of past class failures

```{r}
table_1 <- table(high_use = alc$high_use, number_of_past_class_failures = alc$failures)

addmargins(table_1)

failurel <- ggplot(alc, aes(x = high_use, y = failures))
failurel + geom_count()

```
From this table and plot-picture you can see that the ones who have no past class failures consume less alcohol. Those ones who consume alcohol more have more class failures. 

# Second we explore the relationship between alcohol consumption and absences

```{r}

g <- ggplot(alc, aes(x = high_use, y = absences))

g2 <- g + geom_boxplot() + ggtitle("High alcohol consumption and absences")

g2

```
From this picture we can see that those ones who consume less acohol they have less school absences. Those onse who consume more alcohol have more school absences. So there might be correlation between these two variables.

# Third we explore the correlation between acohol consumption and family relationships
```{r}
g3 <- ggplot(alc, aes(x = high_use, y = famrel))

g4 <- g3 + geom_boxplot()

g4

```
The students who consume less acohol have higher scores from their family relationships. Those student who consume more alcohol have lower scores from their family relationships. From this you can not say which one causes which. Do the students start using alcohol because of bad family relationships or do family relationships get worse when the student starts using more alcohol.

# Fourth we are exploring the hopes for higher education and alcohol consumption
```{r}
table_2 <- table(high_use = alc$high_use, wants_high_education = alc$higher)

table_2

round(prop.table(table_2) * 100, 1)

```
From these two tables you can see that it looks like those who consume more alcohol, they don't want to have higher education.

# Next wer are going to use logistic regression to statistically explore the relationship between my chosen variables and the binary high/low alcohol consumption variable as the target variable.
```{r}
m <- glm(high_use ~ failures + absences + famrel + higher, data = alc, family = "binomial")
summary(m)

```

```{r}
coef(m)
```
As seen above the p-value is low (<0.05) in failures and absences. In famrel the p-value is close to 0.05. The coefficient in failures and absences is positive (failures 0.44, absences 0.08), meaning that more failures and more absences predict higher alcohol use. Familyrelationship seems to have negative effect (-0.22) on the high alcohol use, which means that better familyrelationship have protective effect on alcohol use. The future education plan meaning the plans to get a higher education have also negative effect on the high alcohol use (-0.4), but the p-value is not significant.

# Next we are trying to present and interpret the coefficients of the model as odds ratios and provide confidence intervals for them. 

```{r}
library(tidyverse)
library(dplyr)
```
```{r}
OR <- coef(m) %>% exp()
CI <- exp(confint(m))
```
```{r}
cbind(OR, CI)

```
From this table you can see the coefficients of the model as oddss ratios and their confidence intervals. The odds ratios for failures is 1.54 (and confidence interval is  1.05-2.28).  Variable absences odds ratio is 1.09 (with confidence interval 1.04-1.14). This means that higher consumption of alcohol increases the odds for absence and failures by 1.54 and 1.08 times. The odss ratio for the better family relationship is 0.8 with confidence interval 0.6 - 1.02, which means that the better family relationship makes the odds for high alcohol consumption 0.8 times less likely. The fourth tested variable the higher educational plan was not statistically significant since the odds ratio is 0.70 and the confidence interval is 0.23-2.12. When a number 1 is included in the confidence interval, this means that there is no statistically significant difference. The findings we see can be said to support the earlier hypotheses we had.

# Exploring the predictive power of the model
We are going to use the variables which, according to my logistic regression model, had a statistical relationship with high/low alcohol consumption. We are going to explore the predictive power of my model.
```{r}
m <- glm(high_use ~ failures + absences + famrel + higher, data = alc, family = "binomial")

probabilities <- predict(m, type = "response")

alc <- mutate(alc, probability = probabilities)
alc <- mutate(alc, prediction = probability > 0.5)
select(alc, failures, absences, famrel, high_use, probability, prediction) %>% tail(10)

```
```{r}
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins()

```
```{r}

g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction))

g + geom_point()

```

# The training error
```{r}
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

loss_func(class = alc$high_use, prob = alc$probability)

```
The training errors result is 0.30 meaning 30%. This means that almost 30% a.k.a 1/3 of observations are incorrectly classified. I do not know if this is very common when doing a model. I think the number is quite high and the model is not working as it should be working. 

# Perform 10-fold cross validation on this model
```{r}
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) >0.5
  mean(n_wrong)
}
loss_func(class = alc$high_use, prob = alc$probability)

library(boot)

cv <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = 10)

cv$delta[1]
```
Even after 10-fold cross validation the result remains the same. So 30% of observations are incorrectly classified.