---
author: "Hanna-Leena Kukkola"
title: "chapter6.Rmd"
output: html_document
---
## Week 6
This week we are going to do analysis for longitudinal data. This week was extremely hard for me. Lots of problems with the analysis. I couldn't get all the analysis done correctly. Not all the pictures came out correctly and I had huge problmes with interpreting the results, especially on the BPRS data. Hope to get some points for effort.

First we are going to read the BPRS and RATS datasets.

```{r}
BPRS <- read.csv("BPRS.csv", header = TRUE, sep = ",")
```

```{r}
RATS <- read.csv("RATS.csv", header = TRUE, sep = ",")
```

## We are going to factor the categorial variables again.
```{r}
library(dplyr)
library(tidyr)
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

```

##We need to convert the datasets into long form and we are going to add the time variable to RATS data.
```{r}
#Next step: Convert the data sets to long form an add a Time variable to RATS.

RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4)))
```

Let's draw a table of these datasets. 

```{r}
table(RATSL)
```
This table is quite hard to read because it is so long. But you can see from the table the ID's of the study subjects and you can see that there are three groups that are studied. The dataset of RATS comprised of three different rats group receiving different diets. The weight of these rats were measured in different weekdays. 

```{r}
names(RATS)
```
Whit the command names you can take a look of the variables of the data. These are the variable names of the RATS data and below you can see the variable names of the RATSL data which is the data of the RATS converted to long form and added the variable Time.

```{r}
names(RATSL)
```


##Implement the analyses of Chapter 8 of MABS using the RATS data. First we are going to do analysis for the RATS data.
To begin we shall plot the RATS values, differentiating between the treatment groups into which the rats have been randomized. This simple graph makes a number of features of the data readily apparent. 

```{r}
library(ggplot2)
```

```{r}
# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")

ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

There are 3 different groups studied, each group receiving different diet. From this figure you can see the development of the weight of these rat groups during the almost 70 days period. You can see the individual response profiles by each rat in different groups. As you can see the weight profile in each group is quite different. Group 1 is very different from groups 2 and 3. The group 1 starting weight is also lower than in groups 2 and 3.

### Next we are going to standardize the variables.
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdrats = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()
```
Now we have standardized the data by the variable weight.

##Let's glimpse at the data

```{r}
glimpse(RATSL)
```
Now you can see the standardised variable here in the data. So we succeeded on what we were trying to do.

Let's draw the plot again with the standardised variable.

```{r}
ggplot(RATSL, aes(x = Time, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "stdrats")
```

Now here are the figures with standardised values of the weight. The group 1 is very different from the groups 2 and 3. These rats starting weight was clearly lower than in groups 2 and 3 and the weight gain doesn't change the standardized weight of these rats.

```{r}
head(RATS)
head(RATSL)
```
I used the head command just to look at these two datasets again.

#Now were are going to look at the number of weekdays

```{r}
n <- RATSL$Time %>% unique() %>% length()
```

#Next we are going to look at the summary data with mean and standard error of weight by group and time

```{r}
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()
```

```{r}
glimpse(RATSS)
```

##Next we are going to plot the mean profiles
```{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean -se, ymax = mean +se, linetype = "1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
I personally like this picture very much. It is clear and easy to follow. As you can see there are 3 groups and you can see the individuals in these 3 groups. Each group represents the diet the rats have received. As you can see the group 1 is very different from group 2 and 3. The weight gain is also better in groups 2 and 3.

#Create a summary data by weight and ID with mean as the summary variable (ignoring baseline weekday 0).

```{r}
RATSL8S <- RATSL %>%
  filter(Time>1) %>%
  group_by(ID, Group) %>%
  summarise(mean = mean(Weight)) %>%
  ungroup() 
```

```{r}
glimpse(RATSL8S)
```



#Draw a boxplot 
```{r}
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight, weekdays 2 - 64")
```
From this picture you can see that each group contains individuals who are fare from the box. They are so called outliers and these kind of observations are usually taken away from the analysis so that they don't alter the results too much.

#Create a new data by filtering the outliers and adjust the ggplot code the draw the plot again with the new data.
Next I am going to take the outliers away from the analysis and then draw the picture again.
```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550, mean > 250, mean != 495.2)
```

Drawing the picture again

```{r}
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight, weekdays 2 - 64")
```
As you can see, now the outliers are gone and the boxes look like this. Again you can see that group 1 differs quite a lot from group 2 and 3.

# Perform a two-sample t-test.
In the datacamp we did an exercise where we used the two-sample t-test to the BPRSL data. As you can see from below we cannot do the t-test to the RATSL data because it has 3 different "treatment" groups.
```{r}
names(RATSL8S1)
```
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

#Next I tried to do the t-test to the data, but it was impossible, because there are 3 groups and t-test is used only when comparing  2 groups means to each other to see is there any difference or not.
If there would be only 2 groups, you could use the command as below.

t.test(mean, Weight, data = RATSL8S1)


# Next were are going to add the baseline from the original data as a new variable to the summary data.
```{r}
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)
```


# We are going to fit the linear model with the mean as the response.

```{r}
names(RATSL8S2)
```

Now let's do the fit linear model.
```{r}
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit)
```
As you can see the baseline group has statistical significant difference compared to the group 2 and 3. No statistical difference can be seen in groups 2 and 3. 

#Next we are going to compute the analysis of variance table for the fitted model with anova().
```{r}
anova(fit)
```
This analysis shows that there is a statistical difference between these groups based on the starting weight of these rats. So the baseline group also known as the starting weight of the rats is statistically significant. It is quite understandable, so if the starting weight is higher the rat will weigh more in the end of the study. So it can be so that the starting weight affects more to the end weight than the diet these rats received.

##Then we are going to do next analysis to the BPRS data. Remember that we have to first change the factor variables and then add the week variable to the dataset and also convert the dataset into a long form.
##Impelement the analyses of Chapter 9 of MABS using the BPRS data.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

First let's read the dataset.
```{r}
BPRS <- read.csv("BPRS.csv", header = TRUE, sep = ",")
```

Then we are going to change the factor variables.
```{r}
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
#And then we are going to add a week variable to BPRS and convert it to a long form.
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
```

I didn't know for sure do you need to extract the week number from the dataset but I did it because it was done in the exercises in the datacamp.
```{r}
# Extract the week number
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))
```

Let's look at the BPRSL data. You can use the glimpse and names or heads command to look at the dataset. Also commands view and str can be used for this function.

```{r}
glimpse(BPRSL)
names(BPRSL)
```
As you can see the BPRSL dataset has 360 observations of 5 variables. The end lines shows the names of the variables.

I also did the number of weeks, baseline (week 0) included, bacause it was done in the datacamp exercise.
```{R}
n <- BPRSL$week %>% unique() %>% length()
```

I used the commands names and glimpse because I wanted to look at the BPRSl dataset.
```{r}
names(BPRSL)
glimpse(BPRSL)
```



###Next we are going to draw a plot of the data.
I first draw the plot of the data as it was done in the datacamp.
```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```
The BPRS data consisted of 40 male subjects randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven(extremely severe). As you can see from the picture there are 2 treatment groups. The study subjects were treated for 8 weeks, and you can see how the brps evolves in each subject during these 8 weeks. For me it is hard to see is there a clear difference between these 2 treatment groups because for the majority the bprs is dropping in these 8 weeks in both treatment groups.


First I tried to do the picture with this code below.

ggplot(BPRSL, aes(x = weeks, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) + geom_point(aes(weeks, bprs, color = subject))
  scale_x_continuous(name = "Weeks", breaks = seq(0, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")


I first tried to draw the picture like it was done with the RATS data. Unfortunately I always got this error message. "Error: geom_path: If you are using dotted or dashed lines, colour, size and linetype must be constant over the line". Then I tried to find a solution for this error and I ended up using this code I found geom_step. After that I could draw the picture, but unfortunately I cannot interpret it.

```{r}
ggplot(data = BPRSL) +
  geom_step(aes(x = week, y = bprs, group = subject, color = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```


## Next we are going to create a regression model of the data called BPRS_reg.
First I wanted to look at the variable names of BPRS data and BPRSL data, so I used the commands names.
```{r}
names(BPRS)
```
```{r}
names(BPRSL)
```

To do the next analysis we first need to enter fwe libraries as seen below.
```{r}
library(tidyr)
library(dplyr)
library(lme4)
```

Now we are going to do the regression model of the data.
```{r}
BPRS_reg <- lm(bprs ~ weeks + treatment, data = BPRSL)
```

# Then we are going to print out a summary of the model.
```{r}
summary(BPRS_reg)
```
From the summary table you can see that the p-values are from week2 on very low. Also the higher the multiple R-squared, the higher the multiple R-squared, and the higher F-statistic are the better this model works. So it could be said that the treatment between the tretment weeks affects the bprs results statistically significantly.

#Next we are going to create a random intercept model. So that we can create a random intercept model, we need to access library lme.
```{r}
library(lme4)
```

#Next we are going to create a random intercept model called BPRS_ref.
```{r}
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject  ), data = BPRSL, REML = FALSE)
```

#Next we are going to print out the summary of the model.
```{r}
summary(BPRS_ref)
```
Let's look at the summary table together. From here we can get an estimate of the variance explained by the random effect. This number is important, because if it's indistinguishable from zero, then the random effect probably doesn't matter and we can go ahead and do a regular linear model instead. Next we have estimates of the fixed effects, with standard errors and the t-values. And at last we have the correlation of fixed effects. 

# Next we are going to create a random intercept and random slope model.
```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
```

# Next we are going to print a summary of the model.
```{r}
summary(BPRS_ref1)
```
I am not sure how to interpet the t-values, because they are not the same as p-values. I don't know for sure how to look at the statistical meaning of this summary table, so I am not going to try.

# Next we are going to perform an ANOVA test on the two models.
```{r}
anova(BPRS_ref1, BPRS_ref)
```
The Anova function does a Wald test, which tells us how confident we are of our estimate of the effect of treatment to bprs during 8 study weeks. If I interpret this correctly, the Pr(<Chisq) is low 0.02636*, so does this mean there is statistical significance. 

# Next we are going to create a random intercept and random slope model.
```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week| subject), data = BPRSL, REML = FALSE)
```

# Next we are going to print a summary of the model.
```{r}
summary(BPRS_ref2)
```
I don't know how to interpret this table.

# Next we are going to perform an ANOVA test on the two models.
```{r}
anova(BPRS_ref2, BPRS_ref1)
```
When looking at the comparison of these two models, I think no statistical significant difference is seen, becauset the Pr(>Chisq) is 0.07495, if I interpret this correctly.

# Next we are going to draw the plot of BPRSL.
Again the same error occurs. Error: geom_path: If you are using dotted or dashed lines, colour, size and linetype must be constant over the line.

ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Weeks", breaks = seq(0, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
I tried the same code again. Not even trying to use a different code.

# Next we are going to create a vector of the fitted values.
```{r}
Fitted <- fitted(BPRS_ref2)
```

# Next we are going to create a new column fitted to BPRSL.
```{r}
BPRSL <- BPRSL %>%
  mutate(Fitted)
```

# Next we are going to draw the plot of BPRSL.
I draw the picture of the fitted data using the same code I did before so that the error-message wouldn't come again. I can draw the picture with this geom_step command, but I cannot interpret the results. It all looks very confusing to me.
```{r}

ggplot(data = BPRSL) +
  geom_step(aes(x = week, y = Fitted, group = subject, color = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```

At first I tried the same code as I tried to used before, but the same error occured. Then I changed the code to geom_step, after that I could draw the picture but i don't know how to explain it.

##So this is the end of my results. I had hughe difficulties this week and I think I did a better job with the RATS data than with this BPRS data, but at least I tried. Hope that the peer reviewers can understand some of this.


```{r}
library(magrittr)
```

